### 2020-12-18 paper/sex_selection_spacing-ver5.tex
Tone down sex selection claims

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: The birth of a girl or the birth of a boy. 
I apply the model to the birth histories of Hindu women covering 1972--2016 using 
data from the four National Family and Health Surveys (NFHS).

@@

First, birth intervals lengthened over the four decades, and the lengthening was longer, 
the higher the parity, the more educated the woman, and the higher the percentile. 
Women most likely to use sex selection---well-educated women with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated women in rural areas---still follow the 
traditional pattern of short spacing when they have girls, with only some evidence of sex 
selection. 
The likelihood of a very short birth interval changed little.

Second, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still 
10\%--20\% higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

To separate the effects of the introduction of sex selection and the other changes in
India, I first discuss how the lengths of birth intervals have changed in situations where 
sex selection is less used.
The group broadly covers women with no education, regardless of the sex composition of their 
children, and women with any education who have one or two sons already. 
Despite the lower level of sex selection, son preference is still evident with the 
shortest spacing when they have only girls. 
Notably, for those least likely to use sex selection---rural women with no education---the 
difference in birth interval length across sex compositions has grown over time as spacing when 
sons are present has become longer.

@@

These results are in line with prior research showing that falling fertility is associated 
with a lengthening of particularly the longer birth intervals, although why is still an 
unresolved question \citep{Casterline2016}.
The most educated women who already have a son is an exception to this trend; the 
probability of a third birth declined rapidly, but the birth interval lengths changed little. 
These women both have better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}. 
Hence, it is possible that the lengthening of birth intervals with falling fertility arises
because unintended births become a relatively larger proportion of the total number of
births for a given parity.

@@

The 75th percentile changes are the most dramatic, but sex selection also affects the 
25th and median birth interval lengths. 
For the best-educated urban women with two girls, the 25th percentile birth interval
length increased by six months, or 23\%, while the median percentile birth interval length
increased by 15 months (43\%).

Not surprisingly, given these effects of sex selection, the third spell for the 
best-educated women shows the clearest reversal in the spacing pattern; 
the birth intervals with two girls are consistently longer than the intervals with one or 
two boys, no matter the percentile used.
A similar reversal, although more muted, occurred for the third spell for urban women 
with 1--7 years of education and both urban and rural women with 8--11 years of 
education.

Did the predictions of declining use of sex selection come true? 
There is no clear evidence for or against a reversal in the use of sex selection, with 
some cases showing increases in sex ratios between the last two periods, others little 
change, and some a decline. 
The best-educated women are again a good illustration. 
The sex ratio for women with two girls continued to increase over the last two periods, 
but the likelihood of a third birth declined. 
Furthermore, if the first child was a girl, the sex ratio for the second birth dropped 
slightly, as did the probability of having a second birth. 
However, there are also cases where there is no abatement in the increasing use of sex 
selection. 
For example, for rural women with 1--7 years of education, the sex ratios in the 
absence of girls continued to increase while the likelihood of an additional birth remained 
high.

In summary, over the four decades, birth intervals lengthened with improving economic 
conditions and falling fertility. 
These increases are larger with higher parity and higher percentile measure. 
Furthermore, when sex selection is less used, it appears that the women least likely to 
work are also those with the most substantial increases.

Sex selection, however, is behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth interval lengths. 
Over the four decades, the median birth interval length for this group increased by almost 
15 months, 
and the 75th percentile birth interval length increased by a staggering 21 months, most of 
that within a decade.

@@

Consistent with a more substantial bias in the fertility rate when the age of marriage 
and the length of birth intervals increase, the absolute bias is least in the first 
and the last period and highest in the middle two periods.
Hence, the fertility rate declined too fast from the mid-1980s to the century's end.
Only recently, as the rate of increase for the birth interval lengths has slowed, have the 
two fertility measures begun to converge again.
Even with the convergence, the predicted 2005--2016 cohort fertility is still above the 
1992--1993 fertility rate for every group, except urban women with no education. 
Furthermore, for the last period, the predicted cohort fertility remains at least 
10\%--20\% higher than the fertility rate.

@@

The final question I address is whether there is an association between infant mortality 
and increases in birth spacing and sex selection. 
Starting with the sample used for estimating birth spacing, I select children born more 
than 12 months before the survey month. 
I restrict the analyses to parities two and three because of the small number of births 
and deaths for parity four. 
Furthermore, I do not show the results for women with 12 or more years of education 
for the 1972--1984 period because of the small number of women.

The dependent variable is whether the child died within the first 12 months of life.
The main set of explanatory variables consists of dummies for the spacing from
the prior birth.
The birth interval duration dummies cover 12-month periods, starting nine months after the 
prior birth, until the 57-month dummy, which covers until 105 months after the prior birth.
I use dummies for sex of the index child and the sex composition 
of the prior children.
The birth spacing dummies, the sex of index child, and the sex composition dummies
are all interacted.
Because the actual number of abortions is unobserved, the interactions between
the sex composition of prior children and the sex of the index child serve
as proxies for the use of sex selection.
The other explanatory variables are the same as above, and estimations
are done separately by education level and parity.

@@

An important caveat is that the estimations do not 
address potential selection problems.
For example, suppose women who have diﬀiculties conceiving or carrying a pregnancy to 
term also have a higher mortality risk for their offspring. 
In that case, a spurious correlation between long birth spacing and mortality may arise 
\citep{Kozuki2013}. 
Unfortunately, methods to address selection, such as family fixed effects, do not work 
well when the number of births is as low as for better-educated women 
\citep{Kozuki2013,Molitoris2019}.
However, the fixed effects and linear probability results did not 
deviate substantially in prior research.

@@

Very short birth intervals still exhibit a higher mortality risk, although the 
effect declines with education level.
For the best-educated women, the mortality risk is 3\%--4\%,
whereas women with no education still show a risk that is close to 10\%.

@@

Despite the concern that multiple abortions might increase mortality risk by shortening 
the interval between pregnancies, there is no evidence for this effect. 
Suppose sex-selective abortions lead to higher mortality risk. In that case, boys born 
after a girl---the solid lines---should have an increased risk with longer spacing for 
the two highest education groups in the last two periods.
However, there are no apparent consistent differences between these groups and the other 
potential combinations. 
The same holds for the third spell.

% line 3661 in an_infant_mortality.log
The raw numbers for women with the most uneven sex ratio also suggest that even with 
very high use of sex selection, there is no impact on mortality. 
A total of 1,004 women with 12 or more years of education and no boys at the start of 
the third spell in the last period had a third child, of which 685 were boys. 
Of these 685 boys, only six died within the first year of life. 
Half of those who died were born in the 9--32-month interval, and none in the 57-month+ 
interval.

@@

% Background
Over the past four decades, India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained high. 
Simultaneously, economic growth was strong, schooling increased, and the total
fertility rate fell to close to the replacement level.

@@

% What has happened to spacing and why
The most substantial lengthening of birth intervals came from the best-educated women
because of their substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio increased to close to 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the increase in the long intervals came immediately after the introduction of sex 
selection in India.

Some of these increases are so large that we even observed a reversal of the traditional 
spacing pattern for some groups; when there are no sons, we now see the longest, rather 
than the shortest, birth intervals because of sex selection.
The women who are the least likely to use sex selection still show the traditional 
spacing pattern with short spacing in the absence of sons.

@@

Birth intervals also lengthened in cases when sex selection is less likely to be used. 
However, compared to other countries with similar declines in fertility, the median 
spacing increases were smaller at three to six months over the period. 
Most of the median intervals when sex selection is less used are still short at 36 months 
or below. 
Furthermore, many women still have very short birth intervals. 
In many cases, more than 25\% have their next child within 24 months of the previous 
birth.

% [predictions of declining sex selection] 
Despite predictions that the use of sex selection would decline, there is no clear 
evidence of this. 
The original users of sex selection continue to show substantial male-biased sex ratios, 
although there may be some leveling off. 
More concerning, sex selection appears to be spreading to less educated women as their 
fertility is falling.

@@

The increases in spacing make the total fertility rate a more biased measure of cohort 
fertility. 
This bias was most prominent early in the spread of sex selection when the fertility rate 
was up to one child lower than the predicted cohort fertility. 
However, it is still present, with the predicted cohort fertility 10\%--20\% 
higher than the fertility rate. 
At 1.8 children, the best-educated urban women are the only group for whom the predicted 
cohort fertility is below replacement.

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality 
for the child eventually born.

>>>>

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: The birth of a girl or the birth of a boy. 
I apply the model to Hindu women's birth histories covering 1972--2016 using data from the 
four National Family and Health Surveys (NFHS).

@@

First, birth intervals lengthened over the four decades, with the lengthening longer, the
higher the parity, the more educated the woman, and the higher the percentile.

First, birth intervals lengthened over the four decades, with the lengthening longer, the 
higher the parity, the more educated the woman, and the higher the percentile. 
Well-educated women with no sons had the most substantial lengthening of birth intervals 
and the most male-biased sex ratios, both likely arising from sex-selective abortions. 
Consequently, some women with no sons now have \emph{longer} birth intervals than those 
with sons, reversing India's traditional spacing pattern. 
Less-educated women continued with short spacing when they have girls, with only limited 
evidence of male-biased sex ratios.
The likelihood of a very short birth interval changed little across most groups.

Second, the period fertility rate substantially overestimated how fast cohort fertility 
fell in the 1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still 
10\%--20\% higher than the period fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

To separate the effects of the introduction of sex selection and the other changes in
India, I first discuss how the lengths of birth intervals have changed in situations where, 
based on sex ratios, there appears to be less sex selection.
The group broadly covers women with no education, regardless of their children's sex 
composition, and women with any education who have one or two sons already. 
Despite the apparent lower level of sex selection, son preference is still evident with 
the shortest spacing when they have only girls. 
Notably, for those who appear the least likely to use sex selection---rural women with no 
education---the difference in birth interval length across sex compositions has grown over 
time as spacing when sons are present has become longer.

@@

These results are in line with prior research showing that falling fertility is associated 
with a lengthening of particularly the longer birth intervals, although why is still an 
unresolved question \citep{Casterline2016}.
The most educated women who already have a son is an exception to this trend; the 
probability of a third birth declined rapidly, but the birth interval lengths changed little. 
These women both have better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}. 
Hence, it is possible that the lengthening of birth intervals with falling fertility in 
the prior research arises because unintended births become a relatively larger proportion 
of the total number of births for a given parity.

@@

The 75th percentile changes are the most dramatic, but the apparent use of sex selection 
also affects the 25th and median birth interval lengths. 
For the best-educated urban women with two girls, the 25th percentile birth interval
length increased by six months, or 23\%, while the median percentile birth interval length
increased by 15 months (43\%).

Not surprisingly, given these changes, the third spell for the best-educated women shows
the clearest reversal in the spacing pattern; the birth intervals with two girls are
consistently longer than the intervals with one or two boys, no matter the percentile used.
A similar reversal, although more muted, occurred for the third spell for urban women 
with 1--7 years of education and both urban and rural women with 8--11 years of 
education.

Did the predictions of declining use of sex selection come true? 
There is no clear evidence for or against a reversal in the use of sex selection, with 
some cases showing increases in sex ratios between the last two periods, others little 
change, and some a decline. 
The best-educated women are again a good illustration. 
The sex ratio for women with two girls continued to increase over the last two periods, 
but the likelihood of a third birth declined. 
Furthermore, if the first child was a girl, the sex ratio for the second birth dropped 
slightly, as did the probability of having a second birth. 
However, there are also cases where there is no abatement in the increasing use of sex 
selection. 
For example, for rural women with 1--7 years of education, the sex ratios in the absence 
of daughters continued to increase while the likelihood of an additional birth remained 
high.

In summary, over the four decades, birth intervals lengthened with improving economic 
conditions and falling fertility. 
These increases are larger with higher parity and higher percentile measure. 
Furthermore, when there is little evidence of sex selection use, it appears that the women 
least likely to work are also those with the most substantial increases in birth interval 
lengths.

Sex selection, however, appears to be behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth interval lengths. 
Over the four decades, the median birth interval length for this group increased by almost 
15 months, 
and the 75th percentile birth interval length increased by a staggering 21 months, most of 
that within a decade.

@@

Consistent with a more substantial bias in the period fertility rate when the age of marriage 
and the length of birth intervals increase, the absolute bias is least in the first 
and the last period and highest in the middle two periods.
Hence, the fertility rate declined too fast from the mid-1980s to the century's end.
Only recently, as the rate of increase for the birth interval lengths has slowed, have the 
two fertility measures begun to converge again.
Even with the convergence, the predicted 2005--2016 cohort fertility is still above the 
1992--1993 fertility rate for every group, except urban women with no education. 
Furthermore, for the last period, the predicted cohort fertility remains at least 
10\%--20\% higher than the fertility rate.

@@

The final question I address is whether there is an association between infant mortality 
and increases in birth spacing associated with the introduction of sex selection. 
Starting with the sample used for estimating birth spacing, I select children born more 
than 12 months before the survey month. 
I restrict the analyses to parities two and three because of the small number of births 
and deaths at parity four. 
Furthermore, I do not show the results for women with 12 or more years of education 
for the 1972--1984 period because of the small number of women.

The dependent variable is whether the child died within the first 12 months of life.
The main set of explanatory variables consists of dummies for the spacing from
the prior birth.
The birth interval duration dummies cover 12-month periods, starting nine months after the 
prior birth, until the 57-month dummy, which covers until 105 months after the prior birth.
I use dummies for the sex of the index child and the sex composition 
of the prior children.
The birth spacing dummies, the sex of the index child, and the sex composition dummies
are all interacted.
Because the actual number of abortions is unobserved, the interactions between
the sex composition of prior children and the sex of the index child serve
as proxies for the use of sex selection.
The other explanatory variables are the same as above, and estimations
are done separately by education level and parity.

@@

An important caveat is that the estimations do not address potential selection problems.
For example, suppose women who have diﬀiculties conceiving or carrying a pregnancy to 
term also have a higher mortality risk for their offspring. 
In that case, a spurious correlation between long birth spacing and mortality may arise 
\citep{Kozuki2013}. 
Unfortunately, methods to address selection, such as family fixed effects, do not work 
well when the number of births is as low as it is here for better-educated women 
\citep{Kozuki2013,Molitoris2019}.
However, the fixed effects and linear probability results did not 
deviate substantially in prior research.

@@

Very short birth intervals still exhibit a higher mortality risk, although the 
effect declines with education level.
The mortality risk is 3\%--4\% for the best-educated women, whereas women with no 
education still show a risk that is close to 10\%.

@@

Despite the concern that multiple abortions might increase mortality risk by shortening 
the interval between pregnancies, there is no evidence for this effect. 
Suppose repeated sex-selective abortions lead to a higher mortality risk for the child
eventually born. 
In that case, boys born after a girl---the solid lines---should have an increased risk 
with longer spacing for the two highest education groups in the last two periods.
However, there are no apparent consistent differences between these groups and the other 
potential combinations. 
The same holds for the third spell.

% line 3661 in an_infant_mortality.log
The raw numbers for women with the most uneven sex ratio also suggest that even with a 
very high apparent use of sex selection, there is no impact on mortality. 
A total of 1,004 women with 12 or more years of education and no boys at the start of 
the third spell in the last period had a third child, of which 685 were boys. 
Of these 685 boys, only six died within the first year of life. 
Half of those who died were born in the 9--32-month interval, and none died in the 
57-month+ interval.

@@

% Background
Over the past four decades, India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained high. 
Simultaneously, economic growth was strong, schooling increased, and the total fertility 
rate declined to almost replacement level.

@@

% What has happened to spacing and why
The most substantial lengthening of birth intervals came from the best-educated women,
likely because of substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio reached almost 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the increase in the long intervals came immediately after the introduction of sex 
selection in India.

Some of the lengthening in birth intervals are so large that we even observed a reversal of
the traditional spacing pattern for some groups: 
For these groups, when there are no sons, we now see the longest, rather than the shortest, 
birth intervals, consistent with substantial use of sex selection.
The women who are the least likely to use sex selection, as indicated by their sex ratios
at birth, still show the traditional spacing pattern with short spacing in the absence of
sons.

@@

Birth intervals also lengthened in cases when sex selection appears to be less used.
However, compared to other countries with similar declines in fertility, the median 
spacing increases were smaller at three to six months over the period. 
Most of the median intervals when sex ratios are close to normal are still short at 36 
months or below. 
Furthermore, many women still have very short birth intervals.
More than 25\% of women have their next child within 24 months of the previous birth in 
many cases.

% [predictions of declining sex selection] 
Despite predictions that the use of sex selection would decline, there is no clear 
evidence of this. 
The most likely users of sex selection continue to show substantial male-biased sex ratios, 
although there may be some leveling off.
More concerning, the increasingly male-biased sex ratios among less-educated women suggest
that sex selection use appears to be spreading as fertility falls.

@@

The increases in spacing make the total fertility rate a more biased measure of cohort 
fertility. 
This bias was most prominent early in the spread of access to sex selection when the 
fertility rate was up to one child lower than the predicted cohort fertility. 
However, it is still present, with the predicted cohort fertility 10\%--20\% 
higher than the fertility rate. 
At 1.8 children, the best-educated urban women are the only group for whom the predicted 
cohort fertility is below replacement.

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality
for the child eventually born; boys born after long birth intervals to families with
well-educated mothers, only daughters, and high predicted sex ratios, the combination of
which suggests repeated sex-selective abortions, have no higher infant mortality than other
children.
